### 2010-03-09 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: merge outline into paper. (IDesc)

\documentclass{article}

\usepackage{verbatim}
\usepackage{fullpage}

%% Structure
\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
%\newenvironment{structure}{\comment}{\endcomment}

\begin{document}

\begin{structure}
How to read me:
    '<-'  means "because"
    '->'  means "therefore"
    '/>'  means "but"

The spatial tabbing encodes the scope. Things on the same column are
linked by an implicit "and". Things under an argument justify it.

I know, that's fucked up and you hate it. But that's much better to
read that crap than reading my long frenglish sentences a la James
Proust.
\end{structure}

\section{The Type Theory}

\subsection{Base theory}

\begin{structure}
<- Presentation of the formalism
    <- Standard presentation
        -> No novelty here
    <- 3 judgments [equation]
        -> Context validity
        -> Typing judgements
        -> Equality judgements
    <- Invariants [equation]
        -> By induction on derivations
    <- Judgemental equality
        <- Presentation independant of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{structure}

\begin{structure}
<- Context validity [figure]
    <- Not much to be said
\end{structure}

\begin{structure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{structure}

\begin{structure}
<- Judgemental equality [figure]
    <- symmetry, reflexivity, and transitivity
    <- beta-rules for lambda and pair
    <- xi-rule for functions
    -> Agnostic in the notion of equality
        <- Doesn't rely on a ``propositional'' equality
\end{structure}

\begin{structure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalizing
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{structure}

\subsection{Finite sets}

\begin{structure}
<- Motivation
    <- Finite sets could be encoded with Unit and Bool
        /> Hinder the ability to name things
    -> For convenience
        <- Named elements
        <- Referring by name instead of code
        -> Types as coding presentation
            /> Also as coding representation!
\end{structure}

\begin{structure}
<- Implementation [figure]      
    <- Tags
        -> Purely informational token
    <- EnumU
        -> Universe of finite sets
    <- EnumT e
        -> Elements of finite set e
\end{structure}

\begin{structure}
<- Equipment
    <- \spi operator
        <- Equivalent of Pi on finite sets
        <- First argument: (finite) domain
        <- Second argument: for each element of the domain, a co-domain
        -> Inhabitant of \spi: right-nested tuple of solutions
            <- Skip code for space reasons
    <- switch operator
        <- case analyses over x
        <- index into the \spi tuple to retrieve the corresponding result
\end{structure}

\begin{structure}
<- Equivalent to having a function space over finite sets
    /> Made non-obvious by low-level encodings
        <- General issue with codes
             -> Need to provide an attractive presentation to the user
    -> Types seem to obfuscate our reading
        <- Provide ``too much'' information
        /> False impression: information is actually waiting to be used more widely
        -> See next Section
\end{structure}

\subsection{Type propagation}

\begin{structure}
<- Bidirectional type-checking [ref. Turner,Pierce]
    -> Separating type-checking from type synthesis
    <- Type checking: push terms into types
        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
    <- Type inference: pull types out of terms
        <- Example: |x : S l- x :<: S| gives us the type of x
\end{structure}

\begin{structure}
<- Formalization: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirors typing rule of [figure] 
        -> Ignore identitical judgements
\end{structure}

\begin{structure}
<- Type checking [figure]
    <- Push a type in an exprIn
    <- Result in a full term
    -> *Use* the type to build the term!
        -> Domain and co-domain propagation for Pi and Sigma
        -> Translation of 'tags into EnumTs
        -> Translation of ['tags ...] into EnumUs
        -> Finite function space into switch
\end{structure}

\begin{structure}
<- Type synthesis [figure]
    <- Pull a type out of an exprEx
    <- Result in a full term, together with its type
    -> Do *not* need to specify types
        -> Extracting a term from the context
        -> Function application
        -> Projections
\end{structure}

\begin{structure}
<- Summary
    -> Not a novel technique [refs?]
        /> Used as a boilerplate scrapper
    -> Make dealing with codes *practical*
        <- Example: Finite sets/finite function space
        -> We should not restrain ourself in using codes
            <- We know how to present them to the user
-> Will extend this machinery in further sections
\end{structure}

\section{A universe of simple data-types}

\begin{structure}
<- Why starting with simple data-types
    <- For pedagogical purposes
        <- Data-types as we know them every day
        /> Target dependent types
    -> Cut down version of Induction Recursion
        -> Presentation evolves independently as we go to dependent types
\end{structure}

\subsection{The power of $\Sigma$}

\begin{structure}
<- The duality of Sigma
    <- Sigma generalizes sum over arbitrary arities
        -> \Sigma A B == \Sigma_{x : A} B x
    <- Sigma generalizes product to have a dependant second component
        -> \Sigma A B == (x : A) \times (B x)
\end{structure}

\begin{structure}
<- Data-types in the simply-typed world
    -> "sums-of-product"
        <- Sum of constructors
        <- Product of arguments
<- Data-types in the dependently-typed world
    -> "sigmas-of-sigmas"
    /> Need ability to manipulate these sigmas
        -> Define a Code for data-types
        -> Together with a sigma-based Interpretation
\end{structure}

\subsection{A universe of descriptions}

\begin{structure}
<- Introduction to Universe construction
    <- Define a Code
        -> Name objects
    <- Define an Interpretation of codes into the type theory
        -> Give a semantics to objects
    -> Ability to manipulate code
    -> Ability to compute with these objects
\end{structure}

\begin{structure}
<- Justification of the code 
    <- [both figures]: cannot be read separately
    <- Mimic the standard grammar of data-types description
        -> Just as we already know it
        <- '\Sigma for making sigmas-of-sigmas
        <- 'indx for exhibiting the functoriality
            -> For recursive arguments
        <- '1 for end of description
\end{structure}

\begin{structure}
<- Nat
    <- Sum of zero, suc
    <- zero case: done
    <- suc case: leave open and done
    -> NatD Z = 1 + Z
\end{structure}

\begin{structure}
<- List
    <- Sum of nil, cons
    <- nil case: done
    <- cons case: product of X with leave open and done
    -> ListD X Z = 1 + X * Z
\end{structure}

\begin{structure}
<- Tree
    <- sum of leaf, node
    <- leaf case: done
    <- node case: product of X with two leave open and done
    -> TreeD X Z = 1 + X * Z * Z
\end{structure}

\begin{structure}
<- Fictive object [figure 'data Desc']
    -> Must be read as a type signature
    -> See further for its actual implementation
        <- Subject to our levitation exercise
\end{structure}

\subsection{Fix-point}

\begin{structure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as subobjects
\end{structure}

\begin{structure}
<- Extending type propagation
    <- Data-type declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    -> Coded presentation is practical
        <- The user never see a code
\end{structure}

\begin{structure}
<- Elimination on Mu
    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
        /> Not dependent
        -> Cannot express some (which one again?) properties
    -> Develop a dependent induction
        <- Everywhere/All
        <- Induction
    -> *Generic*
    ???
\end{structure}

\section{Levitating the universe of descriptions}

\subsection{Implementing finite sets}

\begin{structure}
<- Recall typing rules of 1st section
    -> Make clear they were just promises
    -> Can be implemented now
        <- Simply List UId
\end{structure}

\begin{structure}
<- Consequences
    -> Type theory doesn't need to be extended with EnumU, NilE, and ConsE
        <- EnumU == Mu EnumUD
        <- NilE, ConsE are just tags
    -> Do not need a specific \spi eliminator
        <- \spi is an instance of the generic eliminator
            <- Code?
    -> Anything else remains the same (switch, EnumT, 0, 1+)
\end{structure}

\begin{structure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of data-types
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of data-types
\end{structure}

\subsection{Implementing descriptions}

\begin{structure}
<- Realizing our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{structure}

\subsubsection{First try}

\begin{structure}
<- A partial implementation
    <- '1 and 'indx are easy
    <- 'sigma is partially doable
        /> lack the ability to do an higher-order inductive call
    -> Show partial code [figure]
\end{structure}

\subsubsection{Second try}

\begin{structure}
<- Extending the universe of description
    -> With higher-order induction
    <- Intuition: index elements in X by H, and go on reading
        -> indx is isomorph to hindx for H = 1
    /> Keep indx
        <- First order!
        -> Extensionally equal to hindx 1
        /> Practically, definitional equality on Sigma/Pi cannot cope with it
    -> Show DescD code
\end{structure}

\subsubsection{Final move}

\begin{structure}
<- Subtlety: translation of [ ... ]
    -> Let us do it manually
        -> Code with problem for the motive of switch
\end{structure}

\begin{structure}
<- The magician trick
    <- Our problem is to give a motive for switch
        /> We perfectly know what it ought to be: \_ -> DescD
    -> Solution: extend the type theory with a special purpose switchD
        -> Only extension required to the type theory!
        -> Hidden away to the user by the syntactic sugar
            -> Sufficient to ensure unavailability as a raw operator
            <- Another instance of type propagation
\end{structure}

\begin{structure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect data-types
        -> Ability to program on data-types
\end{structure}

\subsubsection{Desc, atomically}

\begin{structure}
<- Adding hindx have introduced some duplication
    <- indx == hindx 1
    -> We can factor out commonalities 
        /> Obtain an equivalent presentation
        /> Still embeddable (refer to the Agda model)
\end{structure}

\begin{structure}
<- Give new presentation [figure]
    <- hindx have introduced the notion of function space: 'Pi
    <- hindx and indx are both composed by a binary product and a left open term: 'x and 'id 
    -> Straightforward translation to the new system [equation]
\end{structure}

\subsection{The generic catamorphism}

\begin{structure}
<- Making cata
    <- Present the type signature
    <- Starts with a call to generic induction
        <- induction on Desc!
        /> Show types at hand
        -> Explain how to use inductive hypothesis
    <- Implement the 'replace' function
    -> Dependent-typeless catamorphism 
\end{structure}

\begin{structure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described by a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{structure}

\subsection{The generic Free Monad}

\begin{structure}
!!! EARLIER !!!
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Garantee a ``constructor form''
\end{structure}

\begin{structure}
<- A generic program: the free monad construction
    <- Recall free monad construction in Haskell
        -> Based on a functor F
    <- Note that the free monad construction is itself defined by a functor
        -> Extract it
    <- Encode it in the Desc world [equation]
        <- F is the Desc we start with
        <- The free monad functor is what we have just defined
        <- [\_]* : Desc -> Set -> Desc
           [\_]* D X = 'cons ['var ('sigma X (\_ -> '1))] D
        -> Mu does the fix-point
\end{structure}

\begin{structure}
<- A generic program: monadic substitution [equation]
    <- subst : \forall T X Y. mu ([T]* X) -> (X -> mu ([T]* Y)) -> mu ([T]* Y)
        -> Using Fold
    -> Consequences
        <- We have free monad data-type
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{structure}

\begin{structure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a data-type
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed data-types
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{structure}

@@

@@

@@

@@

@@

@@

@@

\section{Discussion}

\begin{structure}
!!! Need Help !!!
<- Universe stratification
    <- Stratified agda model
        <- Fully stratified
        <- Proof of iso between host and embedding
    ???
\end{structure}

\end{document}

>>>>

@@

\[
\stk{
\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
\]

\[\stk{
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To \Set \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
\DConst{},
                                                                                    \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} ]) \\
                              & \left[\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
                            & \left[\begin{array}{l}
                                  \DConst{\Fin{n}} \\
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                              \end{array}\right]
\end{array}
}\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: n = \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
                        & \left[
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                          \end{array}
                          \right]
\end{array}
}\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
                     &   &             & \left[
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{n}
                                         \end{array}
                                         \right]
\end{array}
}\]

@@

\[\stk{
\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
                                & \left[\begin{array}{l}
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                                  \end{array}\right]
\end{array}
}\]

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\LambdaTD\: \mapsto \FreeMonad{[\LambdaTFreeD]}{\Fin{}}
}\]

@@

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT [ \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} ]) \\
                                  & \left[\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}\right]
\end{array}
}\]

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\IDescD\: I \mapsto \FreeMonad{[\IDescFreeD]}{I}\: I
}\]
